Pre-registration Study 2 \n Who does not trust basic science?

### 2024-04-01 exp_2/preregistration.Rmd
preregistration study 2

---
title: "Pre-registration Study 1 \n Who does not trust basic science?"
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2:
      keep_md: yes
always_allow_html: true
bibliography: ../references.bib
---

@@

```{r, include=FALSE, message=FALSE}
# read in data from power simulation

```

@@

Anti-vax people or climate skepticists are often conceived of as simply anti-science. But would these people deny that water is made of molecules containing one oxygen and two hydrogen atoms, and that its chemical formula therefore is H2O? We suspect that everyone (or nearly everyone), including people who hold-anti science conspiracy beliefs, trusts basic science, at least when it doesn’t conflict with their specific beliefs.

@@

> *We will ask you 11 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.*

Next, participants answer a set of 11 basic science questions, all but two extracted from existing science knowledge questionnaires. After each question, participants will be presented with an answer reflecting the scientific consensus. Participants are asked to choose whether they accept the answer or not, before proceeding to the next question. Figure \@ref(fig:stimulus-example) displays the survey for an example science question. Table \@ref(tab:knowledge) shows all included questions, their scientifically consensual answer, and their source.

(ref:stimulus-example) Example of a science question, the scientific consensus and the corresponding acceptance question.

```{r stimulus-example, echo=FALSE, out.width= "100%", fig.align="left", fig.show="hold", fig.cap="(ref:stimulus-example)"}
knitr::include_graphics("../figures/study1_question_example.png")
```

@@

```{r knowledge}
items <- read_csv("materials/knowledge_items.csv") %>% 
  mutate(id = 1:nrow(.)) %>% 
  select(id, Question, `Scientific consensus`, `Reference(s)`)

# Output the table
kbl(items, booktabs = T, longtable = TRUE,
    caption = "Science knowledge items") %>%
  kable_paper(full_width = T) %>% 
  column_spec(1, width = "1em") %>%
  column_spec(2, width = "10em") %>%
  column_spec(3, width = "10em") %>%
  column_spec(4, width = "15em") 
```

## Conspiracy scales

@@

-   **RQ2: What is the average acceptance of the scientific consensus (2)?**

Similar to RQ1, we will report the average acceptance rate of the scientific consensus. We will look at the pooled average and the distribution of participant averages.

-   **RQ3: What is the relationship between trust in science and, respectively, (1) and (2)?**

We will regress trust in science on the average knowledge and acceptance per participant.

If they correlate highly, we will use the average of the two Wellcome Global Monitor trust questions. If not, we will focus on the "How much do you trust science?" question.

We will run a robustness check on the Pew question.

-   **RQ4: What is the relationship between conspiracy thinking and, respectively, (1) and (2)?**

We will proceed just as for RQ3, but with conspiracy thinking as outcome, instead of trust.

To measure conspiracy thinking, we will use the average score of the Conspiracy Theory Inventory (BCTI) by @pennycookOverconfidentlyConspiratorialConspiracy2022.

As robustness checks, we will run the same analysis separately for the two other conspiracy scales (also averaging across all items).

>>>>

---
title: "Pre-registration Study 2 \n Who does not trust basic science?"
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2:
      keep_md: yes
always_allow_html: true
bibliography: ../references.bib
---

@@

@@

In a first study (https://osf.io/8utsj/) we tested whether participants would accept the scientific consensus on basic science facts. Most participants did, but not for all the questions. In this survey, we make two main adjustments to further investigate rejection of consensus: (i) we add a longer explanation and sources, instead of only stating the consensus. If it was the case that participants did not mistrust the science, but us the experimenters, this should increase acceptance; (ii) every time a participant rejects the scientific consensus, we ask them an open-ended question to explain why they did so. We also remove one science question that we had added ourselves (on where trees mainly draw there materials from to grow; answer: air), because the results were very different compared to other questions, suggesting the question might have been confusing.

@@

> *We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.*

Next, participants answer a set of 10 basic science questions, all but one extracted from existing science knowledge questionnaires. After each question, participants will be presented with an answer reflecting the scientific consensus, along with an explanation which we wrote, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. Participants are asked to choose whether they accept the answer or not, before proceeding to the next question. Figure 1 displays the survey for an example science question. Table 1 shows all included questions, their scientifically consensual answer, the explanations, and their source.

(ref:stimulus-example) Example of a science question, the explanation with references and the corresponding acceptance question.

```{r stimulus-example, echo=FALSE, out.width= "100%", fig.align="center", fig.show="hold", fig.cap="(ref:stimulus-example)"}
knitr::include_graphics("materials/study2_question_example.png")
```

@@

```{r knowledge}
items <- read_csv("materials/knowledge_items_2.csv") %>% 
  mutate(id = 1:nrow(.)) %>% 
  select(id, Question, `Scientific consensus`, `Explanation`)

# Output the table
kbl(items, booktabs = T, longtable = TRUE,
    caption = "Science knowledge items", linesep = "\\addlinespace") %>%
  kable_paper(full_width = T) %>% 
  column_spec(1, width = "1em") %>%
  column_spec(2, width = "9em") %>%
  column_spec(3, width = "10em") %>%
  column_spec(4, width = "20em") 
```
## Conspiracy scales

@@

In light of the previous study, we change RQ2 to:

-  **RQ2: When a participant’s answer does not match the consensus, how often do they change their mind and accept the consensus (2)?**

Similar to RQ1, we will report the average acceptance rate of the scientific consensus, when it was not already known. We will look at the pooled average and the distribution of participant averages.

-  **RQ3: What reasons do participants provide to justify their rejection of the scientific consensus?**

We will conduct a qualitative analysis and, if there are enough answers, attempt to categorize them.

In light of the results to the first study, we formulate the following hypotheses:

-  **H1a: Higher trust in science is associated with more science knowledge.**
-  **H1b: Higher trust in science is associated with more acceptance of the scientific consensus, given the consensus is not already known.**

We will regress trust in science on the average knowledge and acceptance (for those who initially disagreed) per participant. We will standardize all measures such that the regression coefficient yields a the correlation.

We will focus on the “How much do you trust science?” question, but conduct robustness checks on the two other trust questions.

We will also check how well the three trust items correlate.

-  **H2a: Higher conspiracy thinking is associated with more science knowledge.**
-  **H2b: Higher conspiracy thinking is associated with more acceptance of the scientific consensus, given the consensus is not already known.**

We will proceed just as for H1, but with conspiracy thinking as predictor, instead of trust.

To measure conspiracy thinking, we will use the average score of the Conspiracy Theory Inventory (BCTI) by Pennycook, Binnendyk, and Rand (2022). As robustness checks, we will run the same analysis separately for the two other conspiracy scales.
